---
title: "Ottawa COVID Projections"
---

---
site: blogdown:::blogdown_site
title: "Ottawa COVID-19 Projections"
date: "Last updated `r format(Sys.time(), '%d %B, %Y, %X EST.')`"
output:
  html_document:
    includes:
       in_header: GA_script.html
    theme: cosmo
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
_Last updated on `r format(Sys.time(), '%d %B, %Y, %X')` EST._

La version française de ce contenu sera publiée sous peu.

The charts below show observed & projected hospital resource use for COVID-19 patients in the Ottawa area. The projections supplement provincial projections with current, local data on acute hospital care and intensive care unit (ICU) use.  

The projections are created by the team behind [ProjectBigLife.ca](https://www.projectbiglife.ca) at the Ottawa Hospital and the University of Ottawa with input and data from Ottawa Public Health.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(plotly)
ott_projections <- read.csv(file.path(getwd(), "../Data/Ottawa_hospitalization_estimates_COVID.csv"))
ott_observed <- read.csv(file.path(getwd(), "../Data/Observed data/Ottawa_Observed_COVID_Hospital_Use.csv"))
source("../R/estimate_scripts.R")
```

## Peak hospitalization projections
```{r, echo=FALSE, warning=FALSE, message=FALSE}
peak_projections <- read.csv(file.path(getwd(), "../Data/peak_projections.csv"))
peak_projections$intervention <-
  factor(peak_projections$intervention, levels = c("No prevention & control",
                                                   "50% physical distancing", "70% physical distancing"))
library(tidyverse)
library(ggplot2)
t<- as.tbl(peak_projections) %>% mutate(End = lag(peak_hospital_census),
                                     xpos=1:n()-0.5,
                                    Diff=End-peak_hospital_census,
                                     Percent=paste(round(Diff/End*100,0),"%")) %>%
  ggplot(aes(intervention, peak_hospital_census)) + geom_col(fill = c("#D62728", "#1F77B4", "#2CA02C")) +
  labs(title = "Projected peak hospitalization census in Ottawa by scenario", x = "Scenario",
       y = "Census (# of patients)") + geom_text(aes(label = peak_hospital_census_date), vjust = -0.5) +
  theme(panel.background = element_rect(fill = 'white'), plot.title = element_text(hjust = 0.5)) 
t
```

- 50% physical distancing is a 93% reduction in peak hospitalizations from no prevention & control.

- 70% physical distancing is a 99% reduction in peak hospitalizations from no prevention & control.


> **50% physical distancing means that there is a 50% reduction in close contact between people.** See **[more](/more#distancing)** for a description of physical distancing.

## Acute care hospitalization projections

Click or hover over over top right of visualization for downloading, zooming and panning.

### Census projections

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/covid_visualization.R")
hosp_visualization(ott_projections, ott_observed, "census_acute_care", title = "Projected census of acute care COVID-19 hospitalizations in Ottawa", y_label = "Census (# of patients)", annote = TRUE)
```

### Daily projections

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/covid_visualization.R")
hosp_visualization(ott_projections, ott_observed, "new_acute_care", title = "Projected daily acute care COVID-19 hospitalizations in Ottawa", y_label = "New admissions (# of patients)")
```

## ICU projections

### Census projections

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/covid_visualization.R")
hosp_visualization(ott_projections, ott_observed, "census_ICU", title = "Projected census of COVID-19 ICU hospitalizations in Ottawa", y_label = "Census (# of patients)")
```

### Daily projections

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/covid_visualization.R")
hosp_visualization(ott_projections, ott_observed, "new_ICU", title = "Projected daily COVID-19 ICU hospitalizations in Ottawa", y_label = "New admissions (# of patients)")
```

## Ventilator projections

### Census projections

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/covid_visualization.R")
hosp_visualization(ott_projections, ott_observed, "census_vents", title = "Projected census of COVID-19 ventilator use in Ottawa", y_label = "Census (# of patients)")
```

### Daily projections

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/covid_visualization.R")
hosp_visualization(ott_projections, ott_observed, "new_vents", title = "Projected daily COVID-19 ventilator use in Ottawa", y_label = "Newly ventilated (# of patients)")
```

## Death projections

Death projections were made using existing data from Ottawa Public Health and probability data from [Tuite, Fisman, et al.](https://www.medrxiv.org/content/10.1101/2020.03.24.20042705v1) These projections assume all hospital deaths are in the ICU. Model parameters are [here](/model#parameters). 

Only deaths for hospital patients are included in the projections. Long-term care (LTC, also known as nursing homes) and retirement homes are not included in the projections. As of April 3, 2020, 40 to 60% of COVID-19 related deaths occurred in residents of LTC or retirement homes. Many residents in LTC and retirement homes are in their last years of life and have advanced directives saying they (the resident) do not want advanced life support or transfer to the hospital if they have a life threatening illness. 

### Cumulative deaths in hospital

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/covid_visualization.R")
death_visualization(ott_projections, ott_observed, "cumulative_deaths", title = "Projected cumulative number of COVID-19 hospital deaths in Ottawa", y_label = "Deaths")
```

### Daily projections

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../R/covid_visualization.R")
death_visualization(ott_projections, ott_observed, "new_deaths", title = "Projected number of COVID-19 daily hospital deaths in Ottawa", y_label = "Deaths")
```
