---
title: "Prophet Forecast for older age groups between 50 to 80 year old vaccine 1 dose"
author: "Gauri Saran"
date: '2022-05-08'
output: html_document
---

## <b> Forecast Hospital cases using Prophet model using extra regressor variables from vaccine data for those who received 1st dose of vaccine in the older age groups (50 - 80). 3 days of forecast.</b>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(tidyverse)
library(jsonlite)
library(tidyr)
library(tidyr)
library(lubridate)
library(stats)
library(zoo)
library(stringr)
library(data.table)
library(corrr)
library("Hmisc")
library("corrplot")
library(ggpubr)
library(caTools)
library(Metrics)
source("vaccine_data_prep.R")
source("wastewater.R")
```



```{r echo=FALSE}
#load vaccine data 
ottawa_vaccine_data <-
  fromJSON("https://services.arcgis.com/G6F8XLCl5KtAlZ2G/arcgis/rest/services/COVID-19_Vaccine_Coverage_by_Day_by_Age_by_Dose_for_Ottawa_Residents/FeatureServer/0/query?where=1%3D1&outFields=*&outSR=4326&f=json")

vacc_data <- ottawa_vaccine_data$features$attributes
vacc_data <- clean_vacc_data(vacc_data)

```

### Initial Data Cleaning

Vaccine data, Cases data from the hospital, and the waste water signal data has been loaded, cleaned, and then merged into one final dataframe, `final_data`.

```{r echo=FALSE, warning=FALSE, message = FALSE}
#load vaccine data 
ottawa_vaccine_data <-
  fromJSON("https://services.arcgis.com/G6F8XLCl5KtAlZ2G/arcgis/rest/services/COVID-19_Vaccine_Coverage_by_Day_by_Age_by_Dose_for_Ottawa_Residents/FeatureServer/0/query?where=1%3D1&outFields=*&outSR=4326&f=json")

vacc_data <- ottawa_vaccine_data$features$attributes
#Clean the vaccine data
vacc_data <- clean_vacc_data(vacc_data)

#load cases data
cases_data <- read.csv(file.path(getwd(), "OPH_Observed_COVID_Data.csv"))

# Select only the hospital case column observed_census_ICU_p_acute_care from the data
cases_data <- cases_data %>% mutate(date = as.Date(date))
cases_data <- cases_data %>% select(c(date, observed_census_ICU_p_acute_care))

#load waste water data
ww_data <- read.csv("https://raw.githubusercontent.com/Big-Life-Lab/PHESD/main/Wastewater/Ottawa/Data/wastewater_virus.csv")

# Clean the waste water data
ww_clean <- wastewater_prep(ww_data)
ww_clean <- ww_clean %>% select(c(date, N1_N2_avg))

# Merging the waste water, cases and vaccination data
final_data <- vacc_data %>% inner_join(cases_data, by= "date") %>% inner_join(ww_clean, by = "date") %>% arrange(date)

```

The variables have been log transformed.

```{r echo=FALSE, warning=FALSE, message = FALSE, fig.width=12}
final_data <- final_data %>% select(-c(ObjectId))
final_data$observed_census_ICU_p_acute_care = log(final_data$observed_census_ICU_p_acute_care)
final_data$Percent_of_Ottawa_residents_5_1_dose = log1p(final_data$Percent_of_Ottawa_residents_5_1_dose)
final_data$Percent_of_Ottawa_residents_5_1_dose = log1p(final_data$Percent_of_Ottawa_residents_5_1_dose)
final_data$Percent_of_Ottawa_residents_5_2_dose = log1p(final_data$Percent_of_Ottawa_residents_5_2_dose)
final_data$Percent_of_Ottawa_residents_12_1_dose = log1p(final_data$Percent_of_Ottawa_residents_12_1_dose)
final_data$Percent_of_Ottawa_residents_12_2_dose = log1p(final_data$Percent_of_Ottawa_residents_12_2_dose)
final_data$Percent_of_Ottawa_residents_18_1_dose = log1p(final_data$Percent_of_Ottawa_residents_18_1_dose)
final_data$Percent_of_Ottawa_residents_30_1_dose = log1p(final_data$Percent_of_Ottawa_residents_30_1_dose)
final_data$Percent_of_Ottawa_residents_30_2_dose = log1p(final_data$Percent_of_Ottawa_residents_30_2_dose)
final_data$Percent_of_Ottawa_residents_40_1_dose = log1p(final_data$Percent_of_Ottawa_residents_40_1_dose)
final_data$Percent_of_Ottawa_residents_40_2_dose = log1p(final_data$Percent_of_Ottawa_residents_40_2_dose)
final_data$Percent_of_Ottawa_residents_50_1_dose = log1p(final_data$Percent_of_Ottawa_residents_50_1_dose)
final_data$Percent_of_Ottawa_residents_50_2_dose = log1p(final_data$Percent_of_Ottawa_residents_50_2_dose)
final_data$Percent_of_Ottawa_residents_60_1_dose = log1p(final_data$Percent_of_Ottawa_residents_60_1_dose)
final_data$Percent_of_Ottawa_residents_60_2_dose = log1p(final_data$Percent_of_Ottawa_residents_60_2_dose)
final_data$Percent_of_Ottawa_residents_70_1_dose = log1p(final_data$Percent_of_Ottawa_residents_70_1_dose)
final_data$Percent_of_Ottawa_residents_70_2_dose = log1p(final_data$Percent_of_Ottawa_residents_70_2_dose)
final_data$Percent_of_Ottawa_residents_80_1_dose = log1p(final_data$Percent_of_Ottawa_residents_80_1_dose)
final_data$Percent_of_Ottawa_residents_80_2_dose = log1p(final_data$Percent_of_Ottawa_residents_80_2_dose)
final_data$N1_N2_avg = (final_data$N1_N2_avg) * 1000
final_data$N1_N2_avg = log1p(final_data$N1_N2_avg)

```


The response variable `observed_census_ICU_p_acute_care` has been renamed as y and the date has been renamed as ds to fit the prophet model.

```{r echo=FALSE, warning=FALSE, message = FALSE, fig.width=12, fig.height=12}
final_data <- final_data %>% rename(y = observed_census_ICU_p_acute_care, ds = date)
```

The dataset is divided into train and test set. The test set consist of last 3 days of data.

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Diving data into train and test set.
train <- head(final_data, 409)
test <- tail(final_data, 3)

```

Adding other variable as regressors to the model.

```{r echo=FALSE, warning=FALSE, message = FALSE}
library(prophet)
model_cases <- prophet(interval.width = 0.95)
model_cases <- add_regressor(model_cases, 'N1_N2_avg')
model_cases <- add_regressor(model_cases, 'Percent_of_Ottawa_residents_50_1_dose')
model_cases <- add_regressor(model_cases, 'Percent_of_Ottawa_residents_60_1_dose')
model_cases <- add_regressor(model_cases, 'Percent_of_Ottawa_residents_70_1_dose')
model_cases <- add_regressor(model_cases, 'Percent_of_Ottawa_residents_80_1_dose')

```

Fitting the Prophet Model.

```{r echo=FALSE, warning=FALSE, message = FALSE}
model_cases <- fit.prophet(model_cases, train)
```

Forecasting 3 days into future

```{r echo=FALSE, warning=FALSE, message = FALSE}
future <- make_future_dataframe(model_cases, periods = 3)

```


```{r echo=FALSE, warning=FALSE, message = FALSE}
future$N1_N2_avg = final_data$N1_N2_avg
future$Percent_of_Ottawa_residents_50_1_dose = final_data$Percent_of_Ottawa_residents_50_1_dose
future$Percent_of_Ottawa_residents_60_1_dose = final_data$Percent_of_Ottawa_residents_60_1_dose
future$Percent_of_Ottawa_residents_70_1_dose = final_data$Percent_of_Ottawa_residents_70_1_dose
future$Percent_of_Ottawa_residents_80_1_dose = final_data$Percent_of_Ottawa_residents_80_1_dose
```


```{r echo=FALSE, warning=FALSE, message = FALSE}
forecast <- predict(model_cases, future)


```


Checking last 6 day of forecast data

```{r echo=FALSE, warning=FALSE, message = FALSE}
tail(forecast[c('ds','yhat','yhat_lower', 'yhat_upper')])


```

The plot of actual data and predicted data from Prophet forecast. The blue line is predicted data whereas the black dots are actual data.

```{r echo=FALSE, warning=FALSE, message = FALSE}
dyplot.prophet(model_cases, forecast)

```

Root Mean Squared Error on train data:

```{r echo=FALSE, warning=FALSE, message = FALSE}
prediction_prophet <- exp(forecast$yhat)
actual_data <- exp(final_data$y)
date_prophet <- lubridate::ymd(final_data$ds)
results_prophet <- cbind(date_prophet, prediction_prophet, actual_data)
results_prophet <- as.data.frame(results_prophet)
results_prophet$date_prophet <- as.Date(results_prophet$date_prophet)
rmse(results_prophet$prediction_prophet, results_prophet$actual_data)

```


MAPE on train set:

```{r echo=FALSE, warning=FALSE, message = FALSE}
mape(results_prophet$actual_data,results_prophet$prediction_prophet)
```

Standard deviation of the actual data


```{r echo=FALSE, warning=FALSE, message = FALSE}
sd(exp(final_data$y))

```

Plots comparing actual data and predicted data 


```{r echo=FALSE, warning=FALSE, message = FALSE, fig.width=12}
library(plotly)
linear_reg_plot <- ggplot()+
  geom_line(data=results_prophet,aes(y=actual_data,x= date_prophet,colour="Actual values of hospital cases"),size=1 )+
  geom_line(data=results_prophet,aes(y=prediction_prophet,x= date_prophet,colour="Predicted values of hospital cases"),size=1) +
  scale_color_manual(name = "Y series", values = c("Actual values of hospital cases" = "green", "Predicted values of hospital cases" = "lightblue"))

t <- list(
  family = "Times New Roman",
  size = 16
 )

ggplotly(linear_reg_plot) %>%
  layout(title = list(text = "<b>Comparison plot between Actual and predicted values for Prophet model regressed against those who received 2nd dose.</b>", font = t), xaxis = list(title = ""), yaxis = list(title = "<b>Hospital Cases</b>"), width=1100)

```

RMSE on test set

```{r echo=FALSE, warning=FALSE, message = FALSE}
predicted <- tail(forecast, 3)
date_forecast <- tail(forecast$ds, 3)
actual <- test$y
results <- cbind(date_forecast, predicted, actual)
results <- as.data.frame(results)
results$yhat <- exp(results$yhat)
results$actual <- exp(results$actual)


rmse(results$yhat, results$actual)

```
MAPE on test set

```{r echo=FALSE, warning=FALSE, message = FALSE}
mape(results$actual,results$yhat)
```


Plots comparing actual data and predicted data

```{r echo=FALSE, warning=FALSE, message = FALSE}
library(plotly)
linear_reg_plot <- ggplot()+
  geom_line(data=results,aes(y=actual,x= date_forecast,colour="Actual values of hospital cases"),size=1 )+
  geom_line(data=results,aes(y=yhat,x= date_forecast,colour="Predicted values of hospital cases"),size=1) +
  scale_color_manual(name = "Y series", values = c("Actual values of hospital cases" = "green", "Predicted values of hospital cases" = "red"))

t <- list(
  family = "Times New Roman",
  size = 16
 )

ggplotly(linear_reg_plot) %>%
  layout(title = list(text = "<b>Comparison plot between Actual and predicted values for Linear Regression model</b>", font = t), xaxis = list(title = ""), yaxis = list(title = "<b>Hospital Cases</b>"), width=1100)

```


The error metrics are low when model is regressed against only those who received 1st dose in the older age groups. 


#### <b>Component plot showing yearly and weekly trend of the model.</b>
The model trend shows a strong linear decrease in cases until July 2021 and then it shows a weak to moderate linear increases until January 2022.
The extra regressor plot shows the additive effect of regressors and it shows that 1st dose of vaccination in older age groups shows a strong linear increase until April/May 2021 and then remains steady after July 2021 until January 2022.
Weekly trend shows there are more hospitalizations on Tuesday and Thursday.


```{r echo=FALSE, warning=FALSE, message = FALSE}
prophet_plot_components(model_cases, forecast)


```