---
title: "Generating hospital projections"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This file provides the steps needed to generate covid-19 hospitalization projections. This is based on the [Bayesian extension to the CHIME model](https://github.com/pennsignals/chime_sims) developed by the University of Pennsylvania Health System.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# load libraries and localization files
library(tidyverse)
library(dplyr)
library(jsonlite)
library(reticulate)

source("../R/open_ottawa_scripts.R")
source("../R/chime_functions.R")
```

## Generate data file for simulation
```{r, echo=FALSE, warning=FALSE, message=FALSE}
hosp_export(
  json_address = "https://opendata.arcgis.com/datasets/6bfe7832017546e5b30c5cc6a201091b_0/FeatureServer/0/query?where=1%3D1&outFields=_Date,Cases_Currently_in_Hospital&outSR=4326&f=json", # address for open Ottawa hosp data
  starting_date = "2020-07-07", # starting date of simulation
  dest = "data/OTT_ts.csv")
```

## Running Python simulation
```{r, echo=FALSE, warning=FALSE, message=FALSE}
py_run_file("proj_sim.py")
```

## Integrate new projections with existing projections and export for visualization
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Projection data
old_proj <- read.csv(file.path(getwd(),"../Data/OPH_projection_estimates.csv"))
new_proj <- read.csv(file.path(getwd(),"output/output/hosp_projections.csv"))

# Hospital census & daily admissions from open Ottawa
hosp_object <- fromJSON("https://opendata.arcgis.com/datasets/6bfe7832017546e5b30c5cc6a201091b_0/FeatureServer/0/query?where=1%3D1&outFields=Cases_Newly_Admitted_to_Hospital,Cases_Currently_in_Hospital,_Date&outSR=4326&f=json")
hosp_data <- data_extract(hosp_object)

# Integrate new projections into old projections
hosp_proj <- hosp_integrate(
  old_proj = old_proj, # old projections
  new_proj = new_proj, # newly generated simulation
  starting_date = "2020-07-07" # starting date of simulation
)

# Add hospital data to projections
hosp_admits <- c(hosp_data$Cases_Newly_Admitted_to_Hospital, rep(NA, (nrow(hosp_proj) - nrow(hosp_data))))
hosp_census <- c(hosp_data$Cases_Currently_in_Hospital, rep(NA, (nrow(hosp_proj) - nrow(hosp_data))))

hosp_proj$hosp_census_observed <- hosp_census
hosp_proj$hosp_admits_observed <- hosp_admits

# Export projection for visualization
write.csv(hosp_proj, "../Data/OPH_projection_estimates.csv")
```