---
title: "Ottawa COVID-19 Case Projections"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# libraries
library(plotly)
library(tidyverse)
library(jsonlite)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(stats)
library(covidseir)

# scripts
source("../../R/lag_replacement.R")
source("../../R/observed_data.R")
source("../../R/data_conversion.R")
source("../../R/hosp_projections.R")
source("../../R/open_ottawa_scripts.R")
source("../../R/plot_case_projections.R")

# Pull JSON objects from OpenOttawa
ottawa_case_data <-
  fromJSON("https://opendata.arcgis.com/datasets/6bfe7832017546e5b30c5cc6a201091b_0/FeatureServer/0/query?where=1%3D1&outFields=Cumulative_Active_Cases_by_Episode_Date,7-day_Average_of_Newly_Reported_cases_by_Reported_Date,Daily_Cases_Linked_to_a_Community_Outbreak_by_Episode_Date,Daily_Cases_Not_Linked_to_an_Outbreak_ie_Sporadic_Cases_by_Episode_Date,Daily_Cases_Linked_to_an_Institutional_Outbreak_by_Episode_Date,Cumulative_Deaths_by_Date_of_Death,_Date,Daily_Cases_by_Reported_Date,Daily_Cases_by_Episode_Date,Cases_Newly_Admitted_to_Hospital,Cases_Currently_in_Hospital,Cases_Currently_in_ICU&outSR=4326&f=json")
ottawa_test_data <-
  fromJSON("https://opendata.arcgis.com/datasets/26c902bf1da44d3d90b099392b544b81_0/FeatureServer/0/query?where=1%3D1&outFields=_Date,Number_of_Tests,Daily_%_Positivity&outSR=4326&f=json")

# data
ott_observed_covid <- data_creation(ottawa_case_data, ottawa_test_data)[["data"]]
ott_projections <- read.csv(file.path(getwd(), "../../Data/OPH_projection_estimates.csv"))

# data for dashboard and figures
## Observed census
exp_census <- calc_expected_values_for_n_weeks(data = ott_observed_covid, number_weeks = 3)
## Dt = doubling (or halving) time
Dt <- exp_census[[2]][[3]]
Dt_label = switch( 
  (sign(Dt) + 2),
  "halving time",
  "no growth (flat)",
  "doubling time")
more_less = switch(
  (sign(Dt) + 2),
  "Fewer",
  "About the same",
  "More"
)

# Colors
acute_col <- "rgb(72, 137, 194)"
acute_shade <- "rgba(72, 137, 194, 0.3)"
acute_hex <- "#488AC2"

vent_col <- "rgb(23, 63, 95)"
vent_shade <- "rgba(23, 63, 95, 0.3)"

icu_col <- "rgb(60, 174, 163)"
icu_shade <- "rgba(60, 174, 163)"

current_col <- "rgb(72, 137, 194)"
current_shade <- "rgba(72, 137, 194, 0.3)"
current_hex <- "#488AC2"

reduction_col <- "rgb(243, 213, 86)"
reduction_shade <- "rgba(243, 213, 86, 0.3)"
reduction_hex <- "#F6D556"

increase_col <- "rgb(133, 194, 43)"
increase_shade <- "rgba(133, 194, 43)"
increase_hex <- "#85c22b"

death_col <- "rgb(0,0,0)"
death_hex <- "#000000"
```

# Dashboard

Data to `r as.character(tail(ott_observed_covid$date, n=1))`

### Current

**`r last(na.omit(ott_observed_covid$observed_active_cases))`** active cases in Ottawa.

**`r last(na.omit(ott_observed_covid$observed_census_ICU_p_acute_care))`** confirmed COVID-19 patients in hospital.

Including: **`r last(na.omit(ott_observed_covid$observed_census_acute_care))`** acute care patients. **`r last(na.omit(ott_observed_covid$observed_census_ICU))`** ICU patients.

## Definitions

> **Hospital census** is the count of patients in all Ottawa hospitals at midnight on the previous day.

> **Transmission** is the risk that a person becomes infected when they come in contact with a person with COVID-19. 

> **Doubling time** refers to the time it takes for the number of hospitalized patients to double in number. Conversely, **halving time** refers to the time it takes for the number of hospitalized patients to halve in number.

> **Daily growth** is the rate of change in the hospitalization from day-to-day.

> **Physical distancing** is a reduction in close contact between people. 

> **Physical distancing effectiveness** is the reduction in the transmission of COVID-19 between an infected person and someone who has never had COVID-19, compared to the beginning of the pandemic in March 2020.

> **Safe socializing** is ensuring you don't become a link in a COVID-19 transmission chain that ends in a person becoming hospitalized or dying.

> **Safe reopening** is lifting lockdown measures while ensuring a low risk of COVID-19 hospitalization and death. Safe reopening includes safe socializing, safe work and safe businesses - reopening while ensure we all ensure transmissions chains do not lead to hospitalization and death. 

See [More](/more) page for more information about the terms used.

---

## Observed data
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Set up SEIR model
## Create observed data object that cuts off last 7 days
ott_observed_covid_episodes <- as.data.frame(ott_observed_covid[1:(nrow(ott_observed_covid) - 7),])

## Set up sampling fractions
samp_frac <- c(rep(0.14, 13), rep(0.21, 38))
samp_frac <- c(samp_frac, rep(0.37, nrow(ott_observed_covid_episodes) - length(samp_frac)))

## Set up contact rate breakpoints
f_seg <- c(0, rep(1, nrow(ott_observed_covid_episodes) - 1))
day_new_f <- which(ott_observed_covid_episodes$date == ymd("2020-05-19")) # Date of phase 1 reopening
f_seg[seq(day_new_f, length(f_seg))] <- 2
day_ch <- which(ott_observed_covid_episodes$date == ymd("2020-06-12")) # Date of phase 2 reopening
f_seg[seq(day_ch, length(f_seg))] <- 3
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, results=FALSE}
# Fit SEIR model of observed new cases in Ottawa
fit <- fit_seir(
  daily_cases = ott_observed_covid_episodes$observed_new_episodes,
  samp_frac_fixed = samp_frac, 
  f_seg = f_seg,
  i0_prior = c(log(7), 1), # 7 total cases on March 1st, 2020
  e_prior = c(0.85, 0.05), # 85% of residents social distancing
  start_decline_prior = c(log(24), 0.1), # Starting date of closing of non-essential businesses (March 24th, 2020)
  end_decline_prior = c(log(34), 0.1), # 10 day buffer to ramp in social distancing
  f_prior = cbind(c(0.4, 0.5, 0.6), c(0.2, 0.2, 0.2)),
  R0_prior = c(log(2.5), 0.2), # Starting R0 value
  N_pop = 1049486, # Ottawa population
  iter = 500, # number of posterior samples
  fit_type = "optimizing" # for speed only
)
```
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Visualize model fit
proj <- project_seir(fit, iter = 1:50)
tidy_proj <- tidy_seir(proj, resample_y_rep = 20)
first_day <- min(ott_observed_covid_episodes$date)
last_day <- 500 # how many days to create dates for
lut <- tibble(
  day = seq_len(last_day),
  date = seq(first_day, first_day + length(day) - 1, by = "1 day")
)
```
```{r, echo=FALSE, warning=FALSE, message=FALSE}
tidy_proj <- left_join(tidy_proj, lut, by = "day")
case_projection_plot(tidy_proj, obs_dat = ott_observed_covid_episodes, current_col = current_hex,
  value_column = "observed_new_episodes", date_column = "date", title = "Observed daily COVID-19 cases in Ottawa")
```

---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
observed_daily_ICU_p_acute_care <- list(type = "observed_data", y_column = "observed_new_ICU_p_acute_care", name = "Reported # of patients", color = acute_col)



ICU_p_acute_care <-
  reworked_figure(
    xaxis = "date",
    yaxis = list(observed_daily_ICU_p_acute_care),
    titles = c(y = "# of patients", x = "Date", title = "Observed new COVID-19 hospitalizations of Ottawa residents"),
    data = ott_observed_covid
  )

ICU_p_acute_care
```

---

## Projected case data
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Generate projections for SEIR model fit
ott_case_projections <- proj_generation(fit, lut, days_project = 60, day_start_reduction = 7, pct_change = 10)

# Create data set to visualize projections
projection_data <- case_projection_data(ott_case_projections, 10)
```
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Visualize projections
case_projection_plot(projection_data, ott_observed_covid_episodes, current_col = current_hex, reduction_col = reduction_hex, increase_col = increase_hex, value_column = "observed_new_episodes", date_column = "date", title = "Projected daily COVID-19 cases in Ottawa", pct_change = 10)
```

---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../../R/hosp_projections.R")
hosp_fun(ott_projections, "hosp_census", title = "Projected COVID-19 hospital census in Ottawa", y = "Census (# of patients)", current_color = acute_col, current_shade = acute_shade, reduction_value = 10, reduction_color = reduction_col, reduction_shade = reduction_shade, increase_value = 10, increase_color = increase_col, increase_shade = increase_shade, observed_name = "Observed census of patients")
```

---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("../../R/hosp_projections.R")
hosp_fun(ott_projections, "hosp_admits", title = "Projected COVID-19 daily hospital admissions in Ottawa", y = "New admissions (# of patients)", current_color = acute_col, current_shade = acute_shade, reduction_value = 10, reduction_color = reduction_col, reduction_shade = reduction_shade, increase_value = 10, increase_color = increase_col, increase_shade = increase_shade, observed_name = "Observed # of daily admissions")
```

--- 

## Past deaths
```{r, echo=FALSE, warning=FALSE, message=FALSE}
observed_daily_new_deaths <- list(type = "observed_data", y_column = "observed_new_deaths", name = "Reported # of patients", color = death_col)

new_deaths <-
  reworked_figure(
    xaxis = "date",
    yaxis = list(observed_daily_new_deaths),
    titles = c(y = "# of deaths", x = "Date", title = "Observed COVID-19 daily deaths of Ottawa residents"),
    data = ott_observed_covid
  )

new_deaths

```