---
title: "Short-term case projections for Ottawa (development page)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# load library, packages, functions, data
library(EpiNow2)
library(tidyverse)
library(ggplot2)
source("../../R/epinow_functions.R")
ott_covid_data <- read.csv(file.path(getwd(), "../../Data/Observed Data/OPH_Observed_COVID_Data.csv"))
load("../../Data/short_term_forecast.RData")
ott_projections <- ott_short_forecast[[1]]
dt <- ott_short_forecast[[2]][[5]][[1]]
dt_label = switch( 
  (sign(dt) + 2),
  "Case halving time:",
  "No change:",
  "Case doubling time:")
```

## Dashboard 

(part of the main page dashboard)

`r dt_label` **`r dt` days**

## Introduction 

(text will go after observed data)

Short-term projections are helpful tools for local health units to enact immediate change in order to mitigate the spread of covid-19 in the community. Similar to a weather forecast, short-term projections predict changes in the immediate future; while long-term projections predict changes further down in the future. Paradoxically, short-term & long-term projections can contradict one another. This is because short-term projections are more sensitive to immediate changes; while long-term projections require sustained changes in order projections to change. The following plots illustrate short-term projections for Ottawa. These plots include projected cases by episode date, projected cases by reporting date, and projected reproduction number.

## Short-term projections


```{r, echo=FALSE, warning=FALSE, message=FALSE}
# visualize epinow forecast
short_term_plot(
  projections = ott_projections,
  obs_data = ott_covid_data,
  forecast_type = "infections",
  ylab = "New cases",
  title = "Projections for new cases by episode date"
)

short_term_plot(
  projections = ott_projections,
  obs_data = ott_covid_data,
  forecast_type = "reported_cases",
  ylab = "New cases",
  title = "Projections for new cases by reported date"
)

short_term_plot(
  projections = ott_projections,
  obs_data = ott_covid_data,
  forecast_type = "R",
  ylab = "Reproduction number",
  title = "Reproduction number projections"
)
```