---
title: "Short-term case projections for Ottawa (development page)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## run ott_projections() in short_term_projections.R to generate data for plots
# load library, packages, functions, data
library(tidyverse)
library(ggplot2)
library(EpiNow2)
library(stringr)
library(purrr)
library(lubridate)
library(plotly)
library(RColorBrewer)
source("../../R/epinow_functions.R")
source("../../R/wastewater.R")
ott_covid_data <- read.csv(file.path(getwd(), "../../Data/Observed data/OPH_Observed_COVID_Data.csv"))
ott_observed_waste <- read.csv("https://raw.githubusercontent.com/Big-Life-Lab/covid-19-wastewater/main/data/wastewater_virus.csv")

load("../../Data/short_term_forecast.RData")
load("../../Data/short_term_hosp_proj.RData")
load("../../Data/short_term_ww_proj.RData")

ott_projections <- ott_short_forecast[[1]]

dt <- ott_short_forecast[[2]][[5]][[1]]
dt_label = switch( 
  (sign(dt) + 2),
  "Case halving time:",
  "No change:",
  "Case doubling time:")

dt_abs <- abs(dt)

growth_rate <- ott_short_forecast[[2]][[4]][[1]]
```

## Dashboard 

(part of the main page dashboard)

`r dt_label` **`r dt_abs` days**

## Introduction 

(text will go after observed data)

Short-term and long-term covid-19 projections are presented below. See [More](more) for a discussion about model interpretation and [Methods](methods) for how the projections where created. Short-term projections reflect current covid-19 transmission; while long-term projections reflect sustained transmission trends. Short-term and long-term projections can trend in different directions when there are changing patterns of community covid-19 transmission. 

## Short-term case projections

Projections from `r min(ott_projections$date, na.rm = TRUE)` to `r max(ott_projections$date, na.rm = TRUE)`

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# visualize epinow forecast
short_term_plot(
  projections = ott_projections,
  obs_data = ott_covid_data,
  obs_column = "observed_new_episodes",
  forecast_type = "infections",
  ylab = "New cases",
  title = "Projections for new cases by episode date"
)

short_term_plot(
  projections = ott_projections,
  obs_data = ott_covid_data,
  forecast_type = "reported_cases",
  obs_column = "observed_new_cases",
  ylab = "New cases",
  title = "Projections for new cases by reported date"
)

short_term_plot(
  projections = ott_projections,
  obs_data = ott_covid_data,
  obs_column = "observed_new_cases",
  forecast_type = "R",
  ylab = "R",
  title = "Rt projections"
)
```

## Short-term hospitalization projections

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# visualize epinow forecast
short_term_plot(
  projections = hosp_proj,
  obs_data = ott_covid_data,
  obs_column = "observed_census_ICU_p_acute_care",
  forecast_type = "reported_cases",
  ylab = "Hospital census",
  title = "Projections for hospital census"
)
```

## Short-term wastewater projections

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# visualize epinow forecast
short_term_plot(
  projections = ww_proj,
  obs_data = ww_data,
  obs_column = "secondary",
  forecast_type = "reported_cases",
  ylab = "Wastewater signal (µ)",
  title = "Projections for viral signal"
)
```

## Assessing historic projections with observed data
```{r, echo=FALSE, warning=FALSE, message=FALSE}
load("../../Data/hist_case_projections.RData")
load("../../Data/hist_hosp_projections.RData")
source("../../R/observed_data.R")

case_col <- "rgb(226, 127, 88)"
proj_col <- "rgb(0, 128, 128)"
proj_shade <- "rgba(0, 128, 128, 0.2)"
red <- "rgb(255,0,0)"
red_shade <- "rgba(255, 0, 0, 0.2)"
rolling_avg_col <- "rgb(226, 127, 88)"

ott_data <- ott_covid_data %>%
  mutate(date = as.Date(date)) %>%
  filter(date >= (max(as.Date(date)) - 12*7))

plot_hist_case_proj <- hist_case_projections %>%
  filter(variable == "reported_cases")

plot_hist_hosp_proj <- hist_hosp_projections %>%
  filter(variable == "reported_cases")

case_plot_data <- ott_data %>%
  select(date, observed_new_cases) %>%
  full_join(plot_hist_case_proj) %>%
  filter(duplicated(date) == FALSE,
         date > first(date))

hosp_plot_data <- ott_data %>%
  select(date, observed_census_ICU_p_acute_care) %>%
  full_join(plot_hist_hosp_proj) %>%
  filter(duplicated(date) == FALSE,
         date > first(date))

case_plot_data <- case_plot_data[case_plot_data$date <= max(as.Date(ott_data$date)),]

hosp_plot_data <- hosp_plot_data[hosp_plot_data$date <= max(as.Date(ott_data$date)),]

new_case_call <- list(type = "observed_data", y_column = "observed_new_cases", name = "Observed Historic new cases", short_name = "Daily cases", color = case_col, opacity = 0.5)
new_hosp_call <- list(type = "observed_data", y_column = "observed_census_ICU_p_acute_care", name = "Daily reported new cases", short_name = "Hospital census", color = case_col, opacity = 0.5)
proj_call <- list(type = "avg_data", y_column = "median", name = "Previous projected new cases", short_name = "projections", color = proj_col)
```

```{r}
reworked_figure(xaxis = "date", yaxis = list(proj_call, new_case_call), error_bands = TRUE, error_data = c("upper_90", "lower_90"), error_col = proj_shade, titles = c(y = "New cases", x = "Date", title = "Examining historic case projections with case data"), data = case_plot_data, ticks = FALSE)

reworked_figure(xaxis = "date", yaxis = list(proj_call, new_hosp_call), error_bands = TRUE, error_data = c("upper_90", "lower_90"), error_col = proj_shade, titles = c(y = "Hospital census", x = "Date", title = "Examining historic hospital projections with hospital census data"), data = hosp_plot_data, ticks = FALSE)
```