---
title: "Short-term case projections for Ottawa (development page)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## run ott_projections() in short_term_projections.R to generate data for plots
# load library, packages, functions, data
library(tidyverse)
library(ggplot2)
library(EpiNow2)
library(stringr)
library(purrr)
source("../../R/epinow_functions.R")
ott_covid_data <- read.csv(file.path(getwd(), "../../Data/Observed data/OPH_Observed_COVID_Data.csv"))
load("../../Data/short_term_forecast.RData")

ott_projections <- ott_short_forecast[[1]]

dt <- ott_short_forecast[[2]][[5]][[1]]
dt_label = switch( 
  (sign(dt) + 2),
  "Case halving time:",
  "No change:",
  "Case doubling time:")

growth_rate <- ott_short_forecast[[2]][[4]][[1]]
```

## Dashboard 

(part of the main page dashboard)

`r dt_label` **`r dt` days**

## Introduction 

(text will go after observed data)

Short-term and long-term covid-19 projections are presented below. See [More](more) for a discussion about model interpretation and [Methods](methods) for how the projections where created. Short-term projections reflect current covid-19 transmission; while long-term projections reflect sustained transmission trends. Short-term and long-term projections can trend in different directions when there are changing patterns of community covid-19 transmission. 

## Short-term projections

Projections from `r min(ott_projections$date, na.rm = TRUE)` to `r max(ott_projections$date, na.rm = TRUE)`

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# visualize epinow forecast
#short_term_plot(
  #projections = ott_projections,
  #obs_data = ott_covid_data,
  #forecast_type = "infections",
  #ylab = "New cases",
  #title = "Projections for new cases by episode date"
#)

short_term_plot(
  projections = ott_projections,
  obs_data = ott_covid_data,
  forecast_type = "reported_cases",
  ylab = "New cases",
  title = "Projections for new cases by reported date"
)

#short_term_plot(
  #projections = ott_projections,
  #obs_data = ott_covid_data,
  #forecast_type = "R",
  #ylab = "R",
  #title = "Rt projections"
#)
```