---
title: "Short-term hospital projections for Ottawa (development page)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## run ott_projections() in short_term_projections.R to generate data for plots
# load library, packages, functions, data
library(tidyverse)
library(ggplot2)
library(EpiNow2)
library(stringr)
library(purrr)
library(lubridate)
library(plotly)
library(RColorBrewer)
source("../../R/epinow_functions.R")
source("../../R/wastewater.R")
ott_covid_data <- read.csv(file.path(getwd(), "../../Data/Observed data/OPH_Observed_COVID_Data.csv"))

load("../../Data/short_term_forecast.RData")
load("../../Data/short_term_hosp_proj.RData")
load("../../Data/hosp_projections.RData")

ott_projections <- ott_short_forecast[[1]]

dt <- ott_short_forecast[[3]][[5]][[1]]
dt_label = switch( 
  (sign(dt) + 2),
  "Case halving time:",
  "No change:",
  "Case doubling time:")

dt_abs <- abs(dt)

dt_hosp <- hosp_projections[[3]][[5]][[1]]
dt_label_hosp = switch( 
  (sign(dt_hosp) + 2),
  "Hospital halving time:",
  "No change:",
  "Hospital doubling time:")

dt_abs_hosp <- abs(dt_hosp)

growth_rate <- ott_short_forecast[[2]][[4]][[1]]

time <- .POSIXct(Sys.time(), "America/New_York")


```

_Last updated on `r c(format(time, '%d %B, %Y, %X'))`_

Pour le français cliquez [ici](https://613covid.ca/fr/). 

This website has forecasts of hospitalization counts from COVID-19 and wastewater recovery of SARS-CoV-2.

Website is optimized for desktop use. For mobile users, it is recommended to view in landscape mode. 

Click [here](https://www.ottawapublichealth.ca/en/reports-research-and-statistics/daily-covid19-dashboard.aspx) to access the COVID-19 Daily Dashboard created by Ottawa Public Health. The Ottawa Public Health dashboard provides up to date COVID-19 metrics that inform projections presented on this website. 

## Dashboard 

Data to `r last(ott_covid_data$date)`

### Current

`r dt_label_hosp` **`r dt_abs_hosp` days**

**`r last(ott_covid_data$observed_census_ICU_p_acute_care)`** confirmed COVID-19 patients in hospital.

Including **`r last(ott_covid_data$observed_census_acute_care)`** acute care patients. **`r last(ott_covid_data$observed_census_ICU)`** ICU patients.

## Short-term hospitalization projections

(text will go after observed data)

Short-term COVID-19 hospital projections are presented below. See [More](more) for a discussion about model interpretation and [Methods](methods) for how the projections where created. Short-term projections can trend in different directions when there are changing patterns of community COVID-19 transmission. For example, predicted hospital counts can be underestimated during periods of rapid re-opening.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# visualize epinow forecast
short_term_plot(
  projections = hosp_proj,
  obs_data = ott_covid_data,
  obs_column = "observed_census_ICU_p_acute_care",
  forecast_type = "reported_cases",
  ylab = "Hospital census",
  title = "Projections for hospital census"
)
```

## Short-term case projections

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# visualize epinow forecast
short_term_plot(
  projections = ott_projections,
  obs_data = ott_covid_data,
  obs_column = "observed_new_episodes",
  forecast_type = "infections",
  ylab = "New cases",
  title = "Projections for new cases by episode date"
)

short_term_plot(
  projections = ott_projections,
  obs_data = ott_covid_data,
  forecast_type = "reported_cases",
  obs_column = "observed_new_cases",
  ylab = "New cases",
  title = "Projections for new cases by reported date"
)

short_term_plot(
  projections = ott_projections,
  obs_data = ott_covid_data,
  obs_column = "observed_new_cases",
  forecast_type = "R",
  ylab = "R",
  title = "Rt projections"
)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
load("../../Data/hist_case_projections.RData")
load("../../Data/hist_hosp_projections.RData")
source("../../R/observed_data.R")

case_col <- "rgb(226, 127, 88)"
proj_col <- "rgb(0, 128, 128)"
proj_shade <- "rgba(0, 128, 128, 0.2)"
red <- "rgb(255,0,0)"
red_shade <- "rgba(255, 0, 0, 0.2)"
rolling_avg_col <- "rgb(226, 127, 88)"

ott_data <- ott_covid_data %>%
  mutate(date = as.Date(date)) %>%
  filter(date >= (max(as.Date(date)) - 12*7))

plot_hist_case_proj <- hist_case_projections %>%
  filter(variable == "reported_cases")

plot_hist_hosp_proj <- hist_hosp_projections %>%
  filter(variable == "reported_cases")

case_plot_data <- ott_data %>%
  select(date, observed_new_cases) %>%
  full_join(plot_hist_case_proj) %>%
  filter(duplicated(date) == FALSE,
         date > first(date))

hosp_plot_data <- ott_data %>%
  select(date, observed_census_ICU_p_acute_care) %>%
  full_join(plot_hist_hosp_proj) %>%
  filter(duplicated(date) == FALSE,
         date > first(date))

case_plot_data <- case_plot_data[case_plot_data$date <= max(as.Date(ott_data$date)),]

hosp_plot_data <- hosp_plot_data[hosp_plot_data$date <= max(as.Date(ott_data$date)),]

new_case_call <- list(type = "observed_data", y_column = "observed_new_cases", name = "Observed historic new cases", short_name = "Daily cases", color = case_col, opacity = 0.5)
new_hosp_call <- list(type = "observed_data", y_column = "observed_census_ICU_p_acute_care", name = "Observed historic hospital census", short_name = "Hospital census", color = case_col, opacity = 0.5)
proj_case_call <- list(type = "avg_data", y_column = "median", name = "Historic projected new cases", short_name = "projections", color = proj_col, width = 4)
proj_hosp_call <- list(type = "avg_data", y_column = "median", name = "Historic projected hospital census", short_name = "projections", color = proj_col, width = 4)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#reworked_figure(xaxis = "date", yaxis = list(new_case_call, proj_case_call), error_bands = TRUE, error_data = c("upper_90", "lower_90"), error_col = proj_shade, titles = c(y = "New cases", x = "Date", title = "Predictive model performance: historical comparison of projections to observed new cases"), data = case_plot_data, ticks = FALSE)
```

---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#reworked_figure(xaxis = "date", yaxis = list(new_hosp_call, proj_hosp_call), error_bands = TRUE, error_data = c("upper_90", "lower_90"), error_col = proj_shade, titles = c(y = "Hospital census", x = "Date", title = "Predictive model performance: historical comparison of projections to observed hospital census"), data = hosp_plot_data, ticks = FALSE)
```