---
title: "Ottawa covid-19 wastewater surveillance"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# libraries
library(plotly)
library(tidyverse)
library(jsonlite)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(stats)
library(zoo)
library(stringr)

# scripts
source("../../R/lag_replacement.R")
source("../../R/observed_data.R")
source("../../R/data_conversion.R")
source("../../R/hosp_projections.R")
source("../../R/open_ottawa_scripts.R")
source("../../R/wastewater.R")

# Pull JSON objects from OpenOttawa
ottawa_case_data <-
  fromJSON("https://opendata.arcgis.com/datasets/6bfe7832017546e5b30c5cc6a201091b_0/FeatureServer/0/query?where=1%3D1&outFields=_Date,Cumulative_Cases_by_Episode_Date,Cumulative_Active_Cases_by_Episode_Date,Daily_Cases_by_Reported_Date,7-day_Average_of_Newly_Reported_cases_by_Reported_Date,Daily_Cases_by_Episode_Date,Daily_Cases_Linked_to_a_Community_Outbreak_by_Episode_Date,Daily_Cases_Linked_to_an_Institutional_Outbreak_by_Episode_Date,Daily_Cases_Not_Linked_to_an_Outbreak_ie_Sporadic_Cases_by_Episode_Date,Cases_Newly_Admitted_to_Hospital,Cases_Currently_in_Hospital,Cases_Currently_in_ICU,Cumulative_Deaths_by_Date_of_Death&outSR=4326&f=json")
ottawa_test_data <-
  fromJSON("https://opendata.arcgis.com/datasets/26c902bf1da44d3d90b099392b544b81_0/FeatureServer/0/query?where=1%3D1&outFields=_Date,Number_of_Tests,Daily_%_Positivity&outSR=4326&f=json")

# data
ott_observed_covid <- data_creation(ottawa_case_data, ottawa_test_data)[["data"]]
ott_observed_waste <- read.csv("https://raw.githubusercontent.com/Big-Life-Lab/covid-19-wastewater/main/data/wastewater_virus.csv")

# functions to integrate waste & covid data together
waste_clean <- wastewater_prep(ott_observed_waste)
ott_covid_waste <- merge_data(ott_observed_covid, waste_clean) %>%
    filter(date >= "2020-06-02")

# Colours
case_col <- "rgb(0, 128, 128)"
n1_col <- "rgb(23, 63, 95)"
n2_col <- "rgb(60, 174, 163)"
n1_n2_col <- "rgb(237, 85, 59)"
rolling_avg_col <- "rgb(226, 127, 88)"
```
The National Capital Region is one of the first communities in Canada to conduct daily wastewater readings to help inform its community response in the fight against covid-19, thanks to innovative research from the CHEO Research Institute (CHEO RI) and the University of Ottawa (uOttawa).

The figures below were created by researchers at CHEO RI and uOttawa. Learn more about wastewater epidemiology and its role in covid-19 surveillance on Ottawa Public Health's [website](https://www.ottawapublichealth.ca/en/reports-research-and-statistics/Wastewater_COVID-19_Surveillance.aspx). 

The science supporting wastewater surveillance is rapidly evolving. The plots are currently for research only and presented to the public for discussion only. Wastewater and public health scientists have noted the early wastewater covid-19 signal increase in mid-July that preceded Ottawa summer and fall case resurgence.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Initialize calls for each parameter
new_case_call <- list(type = "observed_data", y_column = "observed_new_cases", name = "Daily reported new cases", short_name = "Daily cases", color = case_col)

new_case_5_day_call <- list(type = "observed_data", y_column = "observed_new_cases_5_day", name = "5 day mean in \nreported new cases", short_name = "5 day", color = case_col, opacity = 0.15)

new_case_10_day_call <- list(type = "observed_data", y_column = "observed_new_cases_10_day", name = "10 day mean in \nreported new cases", short_name = "10 day", color = case_col, opacity = 0.15)

new_case_7_day_call <- list(type = "observed_data", y_column = "observed_new_cases_7_day", name = "7 day mean in \nreported new cases", short_name = "7 day", color = case_col, opacity = 0.15)

new_community_case_5_day_call <- list(type = "observed_data", y_column = "observed_new_community_cases_5_day", name = "5 day mean \nin reported new community cases", color = case_col, opacity = 0.15)

new_community_case_10_day_call <- list(type = "observed_data", y_column = "observed_new_community_cases_7_day", name = "10 day mean \nin reported new community cases", color = case_col, opacity = 0.15)

new_community_case_7_day_call <- list(type = "observed_data", y_column = "observed_new_community_cases_10_day", name = "7 day mean \nin reported new community cases", color = case_col, opacity = 0.15)

hosp_call <- list(type = "observed_data", y_column = "observed_census_ICU_p_acute_care", name = "Hospital census", color = case_col)

active_case_call <- list(type = "observed_data", y_column = "observed_active_cases", name = "Active cases", color = case_col)

active_case_5_day_call <- list(type = "observed_data", y_column = "observed_active_cases_5_day", name = "5 day mean \nin active cases", color = case_col, opacity = 0.15)

active_case_7_day_call <- list(type = "observed_data", y_column = "observed_active_cases_7_day", name = "7 day mean \nin active cases", color = case_col, opacity = 0.15)

active_case_10_day_call <- list(type = "observed_data", y_column = "observed_active_cases_10_day", name = "10 day mean \nin active cases", color = case_col, opacity = 0.15)

n1_call <- list(type = "signal_data", y_column = "N1", name = "N1 RNA signal", color = n1_col, yaxis = "y2") 

n2_call <- list(type = "signal_data", y_column = "N2", name = "N2 RNA signal", color = n2_col, yaxis = "y2")

n1_n2_call <- list(type = "avg_data", y_column = "N1_N2_avg", name = "Avg. viral signal", color = n1_n2_col, yaxis = "y2", opacity = 0.15)

n1_n2_10_day_call <- list(type = "avg_data", y_column = "N1_N2_10_day", name = "10-day mean viral signal", short_name = "10 day", color = rolling_avg_col, yaxis = "y2")

n1_n2_5_day_call <- list(type = "avg_data", y_column = "N1_N2_5_day", name = "5-day mean viral signal", short_name = "5 day", color = rolling_avg_col, yaxis = "y2")

n1_n2_7_day_call <- list(type = "avg_data", y_column = "N1_N2_7_day", name = "7-day mean viral signal", short_name = "7 day", color = rolling_avg_col, yaxis = "y2")

viral_roc_call <- list(type = "avg_data", y_column = "viral_roc_daily", name = "Daily rate of change in viral signal", color = rolling_avg_col, yaxis = "y2")

rolling_avg <- list(type = "avg_data", y_column = "rolling_avg", name = "Daily rate of change in viral signal", color = rolling_avg_col, yaxis = "y2")

viral_rolling_avg_call <- list(type = "avg_data", y_column = "avg_viral_roc_5_day", name = "5-day rolling avg. daily \nchange in viral signal", color = rolling_avg_col, yaxis = "y2")

change_N1_N2_5_day_call <- list(type = "avg_data", y_column = "change_N1_N2_5_day", name = "Percent change in rolling avg \nover 5 days", color = rolling_avg_col, yaxis = "y2")

change_N1_N2_10_day_call <- list(type = "avg_data", y_column = "change_N1_N2_5_day", name = "Percent change in rolling avg \nover 10 days", color = rolling_avg_col, yaxis = "y2")
```

# {#ww-visualization}

```{r, echo=FALSE, warning=FALSE, message=FALSE}
reworked_figure(xaxis = "date", yaxis = list(n1_n2_5_day_call, n1_n2_7_day_call, n1_n2_10_day_call), yaxis2 = list(new_case_10_day_call), titles = c(y = "Normalized viral copies*", y2 = "Number of reported new cases", x = "Date", title = "Ottawa covid-19 viral signal in wastewater"), data = ott_covid_waste, yaxis_button = TRUE)
```
* Viral copies data is normalized using a seasonally stable fecal biomarker; pepper mild mottle virus (PMMoV). See [methods](#/model) for more details.

Data to `r as.character(tail(ott_covid_waste$date, n=1))`