---
title: "Ottawa covid-19 wastewater surveillance (development page) Open science is awesome"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# libraries
library(plotly)
library(tidyverse)
library(jsonlite)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(stats)
library(zoo)
library(stringr)

# scripts
source("../../R/lag_replacement.R")
source("../../R/observed_data.R")
source("../../R/data_conversion.R")
source("../../R/hosp_projections.R")
source("../../R/open_ottawa_scripts.R")
source("../../R/wastewater.R")

# Pull JSON objects from OpenOttawa
ottawa_case_data <-
  fromJSON("https://opendata.arcgis.com/datasets/6bfe7832017546e5b30c5cc6a201091b_0/FeatureServer/0/query?where=1%3D1&outFields=_Date,Cumulative_Cases_by_Episode_Date,Cumulative_Active_Cases_by_Episode_Date,Daily_Cases_by_Reported_Date,7-day_Average_of_Newly_Reported_cases_by_Reported_Date,Daily_Cases_by_Episode_Date,Daily_Cases_Linked_to_a_Community_Outbreak_by_Episode_Date,Daily_Cases_Linked_to_an_Institutional_Outbreak_by_Episode_Date,Daily_Cases_Not_Linked_to_an_Outbreak_ie_Sporadic_Cases_by_Episode_Date,Cases_Newly_Admitted_to_Hospital,Cases_Currently_in_Hospital,Cases_Currently_in_ICU,Cumulative_Deaths_by_Date_of_Death&outSR=4326&f=json")
ottawa_test_data <-
  fromJSON("https://opendata.arcgis.com/datasets/26c902bf1da44d3d90b099392b544b81_0/FeatureServer/0/query?where=1%3D1&outFields=_Date,Number_of_Tests,Daily_%_Positivity&outSR=4326&f=json")

# data
ott_observed_covid <- data_creation(ottawa_case_data, ottawa_test_data)[["data"]]
ott_observed_waste <- read.csv("https://raw.githubusercontent.com/Big-Life-Lab/covid-19-wastewater/main/data/wastewater_virus.csv")

# functions to integrate waste & covid data together
waste_clean <- wastewater_prep(ott_observed_waste)
ott_covid_waste <- merge_data(ott_observed_covid, waste_clean) %>%
    filter(date >= "2020-06-02")

# Colours
case_col <- "rgb(0, 128, 128)"
n1_col <- "rgb(23, 63, 95)"
n2_col <- "rgb(60, 174, 163)"
n1_n2_col <- "rgb(237, 85, 59)"
rolling_avg_col <- "rgb(226, 127, 88)"
```
The National Capital Region is the first community in Canada to conduct daily wastewater readings to help inform its community response in the fight against covid-19, thanks to innovative research from the CHEO Research Institute (CHEO RI) and the University of Ottawa (uOttawa).

Learn more about wastewater epidemiology and its role in covid-19 surveillance on Ottawa Public Health's [website](https://www.ottawapublichealth.ca/en/reports-research-and-statistics/Wastewater_COVID-19_Surveillance.aspx). 

The science supporting wastewater surveillance is rapidly evolving. See the [Methods](/model/) page for information on how the samples where collected, access to the data, and how the plots were created. The plots are currently for research only and presented to the public for discussion. Wastewater surveillance is currently viewed as having several uses, including:

- Early identification of new cases of covid-19 when there is a low level of covid-19 infection.
- A population perspective of covid-19 infection that is not influenced by the level of covid-19 testing. Everyone pees and poos, including people with and without covid-19. People with covid-19 shed SARS-CoV-2 in their feces, regardless of whether they receive a covid-19 test or diagnosis. 

However, at this time, we recommend caution when interpretation daily and short-term variation in the viral signal. Covid-19 wastewater signal is helpful when interpreted alongside other covid-19 surveillance measures, taking into considerations of the strengths and limitations of each measure. The accuracy of SARS-CoV-2 testing is improving, as scientists understand the role of factors such differences in sewage systems and laboratory protocols.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Initialize calls for each parameter
new_case_call <- list(type = "observed_data", y_column = "observed_new_cases", name = "Daily reported new cases", short_name = "Daily cases", color = case_col)

new_case_5_day_call <- list(type = "observed_data", y_column = "observed_new_cases_5_day", name = "5-day midpoint mean in \nreported new cases \n(by reporting date)", short_name = "New cases\n(reported date)", color = case_col, opacity = 0.15)

new_case_10_day_call <- list(type = "observed_data", y_column = "observed_new_cases_10_day", name = "10-day midpoint mean in \nreported new cases \n(by reporting date)", short_name = "New cases\n(reported date)", color = case_col, opacity = 0.15)

new_case_7_day_call <- list(type = "observed_data", y_column = "observed_new_cases_7_day", name = "7-day midpoint mean in \nreported new cases \n(by reporting date)", short_name = "New cases\n(reported date)", color = case_col, opacity = 0.15)

new_episode_5_day_call <- list(type = "observed_data", y_column = "observed_new_episodes_5_day", name = "5-day mean in \nreported new cases \n(by episode date)", short_name = "New cases\n(episode date)", color = case_col, opacity = 0.15)

new_episode_10_day_call <- list(type = "observed_data", y_column = "observed_new_episodes_10_day", name = "10-day mean in \nreported new cases \n(by episode date)", short_name = "New cases\n(episode date)", color = case_col, opacity = 0.15)

new_episode_7_day_call <- list(type = "observed_data", y_column = "observed_new_episodes_7_day", name = "7-day mean in \nreported new cases \n(by episode date)", short_name = "New cases\n(episode date)", color = case_col, opacity = 0.15)

new_community_case_5_day_call <- list(type = "observed_data", y_column = "observed_new_community_cases_5_day", name = "5-day mean \nin reported new community cases", color = case_col, opacity = 0.15)

new_community_case_10_day_call <- list(type = "observed_data", y_column = "observed_new_community_cases_7_day", name = "10-day mean \nin reported new community cases", color = case_col, opacity = 0.15)

new_community_case_7_day_call <- list(type = "observed_data", y_column = "observed_new_community_cases_10_day", name = "7 day mean \nin reported new community cases", color = case_col, opacity = 0.15)

hosp_call <- list(type = "observed_data", y_column = "observed_census_ICU_p_acute_care", name = "Hospital census", short_name = "Hosp census", color = case_col, opacity = 0.15)

pct_positivity_call_7_day <- list(type = "observed_data", y_column = "pct_positivity_7_day", name = "7-day mean \npercent positivity rate", short_name = "% positivity", color = case_col, opacity = 0.15)

active_case_call <- list(type = "observed_data", y_column = "observed_active_cases", name = "Active cases", short_name = "Active cases", color = case_col, opacity = 0.15)

active_case_5_day_call <- list(type = "observed_data", y_column = "observed_active_cases_5_day", name = "5-day mean \nin active cases", color = case_col, opacity = 0.15)

active_case_7_day_call <- list(type = "observed_data", y_column = "observed_active_cases_7_day", name = "7-day mean \nin active cases", short_name = "Active cases", color = case_col, opacity = 0.15)

active_case_10_day_call <- list(type = "observed_data", y_column = "observed_active_cases_10_day", name = "10-day mean \nin active cases", color = case_col, opacity = 0.15)

n1_call <- list(type = "signal_data", y_column = "N1", name = "N1 RNA signal", color = n1_col, yaxis = "y2") 

n2_call <- list(type = "signal_data", y_column = "N2", name = "N2 RNA signal", color = n2_col, yaxis = "y2")

n1_n2_call <- list(type = "avg_data", y_column = "N1_N2_avg", name = "Avg. viral signal", short_name = "viral signal", color = n1_n2_col, yaxis = "y2", opacity = 0.15)

n1_n2_10_day_call <- list(type = "avg_data", y_column = "N1_N2_10_day", name = "10-day midpoint mean viral signal", short_name = "10-day", color = rolling_avg_col, yaxis = "y2")

n1_n2_5_day_call <- list(type = "avg_data", y_column = "N1_N2_5_day", name = "5-day midpoint mean viral signal", short_name = "5-day", color = rolling_avg_col, yaxis = "y2")

n1_n2_7_day_call <- list(type = "avg_data", y_column = "N1_N2_7_day", name = "7-day midpoint mean viral signal", short_name = "7-day", color = rolling_avg_col, yaxis = "y2")

viral_roc_call <- list(type = "avg_data", y_column = "viral_roc_daily", name = "Daily rate of change in viral signal", color = rolling_avg_col, yaxis = "y2")

rolling_avg <- list(type = "avg_data", y_column = "rolling_avg", name = "Daily rate of change in viral signal", color = rolling_avg_col, yaxis = "y2")

viral_rolling_avg_call <- list(type = "avg_data", y_column = "avg_viral_roc_5_day", name = "5-day rolling avg. daily \nchange in viral signal", color = rolling_avg_col, yaxis = "y2")

change_N1_N2_5_day_call <- list(type = "avg_data", y_column = "change_N1_N2_5_day", name = "Percent change in rolling avg \nover 5 days", short_name = "5-day % change", color = rolling_avg_col, yaxis = "y2")

change_N1_N2_10_day_call <- list(type = "avg_data", y_column = "change_N1_N2_10_day", name = "Percent change in rolling avg \nover 10 days", short_name = "10-day % change", color = rolling_avg_col, yaxis = "y2")

change_N1_N2_7_day_call <- list(type = "avg_data", y_column = "change_N1_N2_7_day", name = "Percent change in viral signal rolling avg \nover 7 days", short_name = "7 day % change", color = rolling_avg_col, yaxis = "y2")

change_new_cases_5_day_call <- list(type = "observed_data", y_column = "change_new_cases_5_day", name = "Percent change in rolling avg \nover 5 days", short_name = "5 day % change", color = case_col, yaxis = "y2", opacity = 0.15)

change_new_cases_10_day_call <- list(type = "observed_data", y_column = "change_new_cases_10_day", name = "Percent change in rolling avg \nover 10 days", short_name = "10 day % change", color = case_col, yaxis = "y2", opacity = 0.15)

change_new_cases_7_day_call <- list(type = "observed_data", y_column = "change_new_cases_7_day", name = "Percent change in new case rolling avg \nover 7 days", short_name = "7 day % change", color = case_col, yaxis = "y2", opacity = 0.15)

daily_viral_signal_call <- list(type = "observed_data", y_column = "N1_N2_avg", name = "Daily viral signal", short_name = "Daily signal", color = rolling_avg_col, yaxis = "y2", opacity = 0.50)
```

# {#ww-visualization}

```{r, echo=FALSE, warning=FALSE, message=FALSE}
reworked_figure(xaxis = "date", yaxis = list(daily_viral_signal_call, n1_n2_7_day_call), titles = c(y = "Normalized viral copies*", y2 = "", x = "Date", title = "Covid-19 wastewater viral signal, Ottawa"), data = ott_covid_waste)
```

---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
reworked_figure(xaxis = "date", yaxis = list(n1_n2_7_day_call), yaxis2 = list(new_case_7_day_call, new_episode_7_day_call, active_case_call, pct_positivity_call_7_day, hosp_call), titles = c(y = "Normalized viral copies*", y2 = "", x = "Date", title = "7-day mean covid-19 wastewater viral signal, Ottawa"), data = ott_covid_waste, yaxis2_button = TRUE, y2_button_name = "Comparator")
```


---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
reworked_figure(xaxis = "date", yaxis = list(change_N1_N2_7_day_call, change_new_cases_7_day_call), titles = c(y = "Percent change (%)", x = "Date", title = "Percent change^ in 7 day rolling average"), data = ott_covid_waste)
```

7-day averages are rolling averages, calculated using the results for the previous days. The frequency of testing changes over time. Currently, Ottawa wastewater testing is performed 5 days each week.

\* New cases (reported date) is the date reported by the laboratory. New cases (episode date) is the earliest onset, test or reported date.

\* Viral copies data is normalized using a seasonally stable fecal biomarker; pepper mild mottle virus (PMMoV). See [methods](/model/) for more details.

^ Percent change in 7-day average is calculated by comparing the 7-day average (previous day 1 to 7) with a lagged 5-day average (days 8 to 14).

Data to `r as.character(tail(ott_covid_waste$date, n=1))`