---
title: "Ottawa covid-19 wastewater surveillance (development page)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# libraries
library(plotly)
library(tidyverse)
library(jsonlite)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(stats)
library(zoo)
library(stringr)

# scripts
source("../../R/lag_replacement.R")
source("../../R/observed_data.R")
source("../../R/data_conversion.R")
source("../../R/hosp_projections.R")
source("../../R/open_ottawa_scripts.R")
source("../../R/wastewater.R")

# Pull JSON objects from OpenOttawa
ottawa_case_data <-
  fromJSON("https://opendata.arcgis.com/datasets/6bfe7832017546e5b30c5cc6a201091b_0/FeatureServer/0/query?where=1%3D1&outFields=_Date,Cumulative_Cases_by_Episode_Date,Cumulative_Active_Cases_by_Episode_Date,Daily_Cases_by_Reported_Date,7-day_Average_of_Newly_Reported_cases_by_Reported_Date,Daily_Cases_by_Episode_Date,Daily_Cases_Linked_to_a_Community_Outbreak_by_Episode_Date,Daily_Cases_Linked_to_an_Institutional_Outbreak_by_Episode_Date,Daily_Cases_Not_Linked_to_an_Outbreak_ie_Sporadic_Cases_by_Episode_Date,Cases_Newly_Admitted_to_Hospital,Cases_Currently_in_Hospital,Cases_Currently_in_ICU,Cumulative_Deaths_by_Date_of_Death&outSR=4326&f=json")
ottawa_test_data <-
  fromJSON("https://opendata.arcgis.com/datasets/26c902bf1da44d3d90b099392b544b81_0/FeatureServer/0/query?where=1%3D1&outFields=_Date,Number_of_Tests,Daily_%_Positivity&outSR=4326&f=json")

# data
ott_observed_covid <- data_creation(ottawa_case_data, ottawa_test_data)[["data"]]
ott_observed_waste <- read.csv("https://raw.githubusercontent.com/Big-Life-Lab/covid-19-wastewater/main/data/wastewater_virus.csv")

# functions to integrate waste & covid data together
waste_clean <- wastewater_prep(ott_observed_waste)
ott_covid_waste <- merge_data(ott_observed_covid, waste_clean) %>%
    filter(date >= "2020-06-02")

# Colours
case_col <- "rgb(0, 128, 128)"
n1_col <- "rgb(23, 63, 95)"
n2_col <- "rgb(60, 174, 163)"
n1_n2_col <- "rgb(237, 85, 59)"
rolling_avg_col <- "rgb(226, 127, 88)"
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Initialize calls for each parameter
new_case_call <- list(type = "observed_data", y_column = "observed_new_cases", name = "Daily reported new cases", short_name = "Daily cases", color = case_col)

new_case_5_day_call <- list(type = "observed_data", y_column = "observed_new_cases_5_day", name = "5-day midpoint mean in \nreported new cases \n(by reporting date)", short_name = "New cases\n(reported date)", color = case_col, opacity = 0.15)

new_case_10_day_call <- list(type = "observed_data", y_column = "observed_new_cases_10_day", name = "10-day midpoint mean in \nreported new cases \n(by reporting date)", short_name = "New cases\n(reported date)", color = case_col, opacity = 0.15)

new_case_7_day_call <- list(type = "observed_data", y_column = "observed_new_cases_7_day", name = "7-day midpoint mean in \nreported new cases \n(by reporting date)", short_name = "New cases\n(reported date)", color = case_col, opacity = 0.15)

new_episode_5_day_call <- list(type = "observed_data", y_column = "observed_new_episodes_5_day", name = "5-day mean in \nreported new cases \n(by episode date)", short_name = "New cases\n(episode date)", color = case_col, opacity = 0.15)

new_episode_10_day_call <- list(type = "observed_data", y_column = "observed_new_episodes_10_day", name = "10-day mean in \nreported new cases \n(by episode date)", short_name = "New cases\n(episode date)", color = case_col, opacity = 0.15)

new_episode_7_day_call <- list(type = "observed_data", y_column = "observed_new_episodes_7_day", name = "7-day mean in \nreported new cases \n(by episode date)", short_name = "New cases\n(episode date)", color = case_col, opacity = 0.15)

new_community_case_5_day_call <- list(type = "observed_data", y_column = "observed_new_community_cases_5_day", name = "5-day mean \nin reported new community cases", color = case_col, opacity = 0.15)

new_community_case_10_day_call <- list(type = "observed_data", y_column = "observed_new_community_cases_7_day", name = "10-day mean \nin reported new community cases", color = case_col, opacity = 0.15)

new_community_case_7_day_call <- list(type = "observed_data", y_column = "observed_new_community_cases_10_day", name = "7 day mean \nin reported new community cases", color = case_col, opacity = 0.15)

hosp_call <- list(type = "observed_data", y_column = "observed_census_ICU_p_acute_care", name = "Hospital census", short_name = "Hosp census", color = case_col, opacity = 0.15)

pct_positivity_call_7_day <- list(type = "observed_data", y_column = "pct_positivity_7_day", name = "7-day mean \npercent positivity rate", short_name = "% positivity", color = case_col, opacity = 0.15)

active_case_call <- list(type = "observed_data", y_column = "observed_active_cases", name = "Active cases", short_name = "Active cases", color = case_col, opacity = 0.15)

active_case_5_day_call <- list(type = "observed_data", y_column = "observed_active_cases_5_day", name = "5-day mean \nin active cases", color = case_col, opacity = 0.15)

active_case_7_day_call <- list(type = "observed_data", y_column = "observed_active_cases_7_day", name = "7-day mean \nin active cases", short_name = "Active cases", color = case_col, opacity = 0.15)

active_case_10_day_call <- list(type = "observed_data", y_column = "observed_active_cases_10_day", name = "10-day mean \nin active cases", color = case_col, opacity = 0.15)

n1_call <- list(type = "signal_data", y_column = "N1", name = "N1 RNA signal", color = n1_col, yaxis = "y2") 

n2_call <- list(type = "signal_data", y_column = "N2", name = "N2 RNA signal", color = n2_col, yaxis = "y2")

n1_n2_call <- list(type = "avg_data", y_column = "N1_N2_avg", name = "Avg. viral signal", short_name = "viral signal", color = n1_n2_col, yaxis = "y2", opacity = 0.15)

n1_n2_10_day_call <- list(type = "avg_data", y_column = "N1_N2_10_day", name = "10-day midpoint mean viral signal", short_name = "10-day", color = rolling_avg_col, yaxis = "y2")

n1_n2_5_day_call <- list(type = "avg_data", y_column = "N1_N2_5_day", name = "5-day midpoint mean viral signal", short_name = "5-day", color = rolling_avg_col, yaxis = "y2")

n1_n2_7_day_call <- list(type = "avg_data", y_column = "N1_N2_7_day", name = "7-day midpoint mean viral signal", short_name = "7-day", color = rolling_avg_col, yaxis = "y2")

viral_roc_call <- list(type = "avg_data", y_column = "viral_roc_daily", name = "Daily rate of change in viral signal", color = rolling_avg_col, yaxis = "y2")

rolling_avg <- list(type = "avg_data", y_column = "rolling_avg", name = "Daily rate of change in viral signal", color = rolling_avg_col, yaxis = "y2")

viral_rolling_avg_call <- list(type = "avg_data", y_column = "avg_viral_roc_5_day", name = "5-day rolling avg. daily \nchange in viral signal", color = rolling_avg_col, yaxis = "y2")

change_N1_N2_5_day_call <- list(type = "avg_data", y_column = "change_N1_N2_5_day", name = "Percent change in rolling avg \nover 5 days", short_name = "5-day % change", color = rolling_avg_col, yaxis = "y2")

change_N1_N2_10_day_call <- list(type = "avg_data", y_column = "change_N1_N2_10_day", name = "Percent change in rolling avg \nover 10 days", short_name = "10-day % change", color = rolling_avg_col, yaxis = "y2")

change_N1_N2_7_day_call <- list(type = "avg_data", y_column = "change_N1_N2_7_day", name = "Percent change in \nviral signal rolling \navg over 7 days", short_name = "7 day % change", color = rolling_avg_col, yaxis = "y2")

change_new_cases_5_day_call <- list(type = "observed_data", y_column = "change_new_cases_5_day", name = "Percent change in rolling avg \nover 5 days", short_name = "5 day % change", color = case_col, yaxis = "y2", opacity = 0.15)

change_new_cases_10_day_call <- list(type = "observed_data", y_column = "change_new_cases_10_day", name = "Percent change in rolling avg \nover 10 days", short_name = "10 day % change", color = case_col, yaxis = "y2", opacity = 0.15)

change_new_cases_7_day_call <- list(type = "observed_data", y_column = "change_new_cases_7_day", name = "Percent change in \nnew case rolling \navg over 7 days", short_name = "7 day % change", color = case_col, yaxis = "y2", opacity = 0.15)

daily_viral_signal_call <- list(type = "observed_data", y_column = "N1_N2_avg", name = "Daily viral signal", short_name = "Daily signal", color = rolling_avg_col, yaxis = "y2", opacity = 0.50)
```

The National Capital Region is the first community in Canada to conduct daily wastewater readings to help inform its community response in the fight against covid-19, thanks to innovative research from the CHEO Research Institute (CHEO RI) and the University of Ottawa (uOttawa).

Learn more about wastewater epidemiology and its role in covid-19 surveillance on Ottawa Public Health's [website](https://www.ottawapublichealth.ca/en/reports-research-and-statistics/Wastewater_COVID-19_Surveillance.aspx). 

The following plots below illustrate daily viral covid-19 wastewater signal in Ottawa. We recommend caution when interpreting daily and short-term variation in the viral signal. Covid-19 wastewater signal is helpful when interpreted alongside other covid-19 surveillance measures, taking into consideration the strengths and limitations of each measure. The accuracy of SARS-CoV-2 testing is improving, as scientists understand the role of factors such as differences in sewage systems and laboratory protocols.

# {#ww-visualization}

```{r, echo=FALSE, warning=FALSE, message=FALSE}
reworked_figure(xaxis = "date", yaxis = list(daily_viral_signal_call, n1_n2_7_day_call), titles = c(y = "Normalized viral copies*", y2 = "", x = "Date", title = "Covid-19 wastewater viral signal, Ottawa"), data = ott_covid_waste)
```

---

See the [Methods](/model/) page for more information on how the samples where collected, access to the data, and how the plots were created. The plots are currently for research only and presented to the public for discussion. Wastewater surveillance has several uses, including:

- A population perspective of covid-19 infection. Everyone uses the toilet, including people with and without covid-19. People with covid-19 shed SARS-CoV-2 in their feces, regardless of whether they receive a covid-19 test or diagnosis. 
- Early identification of new cases of covid-19 when there is a low level of covid-19 infection.
Wastewater surveillance provides a unique population perspective. For example, reported covid-19 cases underrepresents asymptomatic infection and it is subject to changing testing criteria. The following plot illustrates daily covid-19 wasterwater viral signal along with various comparators including reported daily covid-19 cases and hospitalization census. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
reworked_figure(xaxis = "date", yaxis = list(n1_n2_7_day_call), yaxis2 = list(new_case_7_day_call, new_episode_7_day_call, active_case_call, pct_positivity_call_7_day, hosp_call), titles = c(y = "Normalized viral copies*", y2 = "", x = "Date", title = "7-day mean covid-19 wastewater viral signal, Ottawa"), data = ott_covid_waste, yaxis2_button = TRUE, y2_button_name = "Comparator")
```

---

Early detection of covid-19 outbreaks and surges results in a quick _change_ in the number of people covid-19 infection. The following plot shows the change in covid-19 infection and the potential of wastewater surveillance to identify outbreaks and surges. Note the 580% increase in signal on July 6 that preceded a covid-19 summer surge.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
reworked_figure(xaxis = "date", yaxis = list(change_N1_N2_7_day_call, change_new_cases_7_day_call), titles = c(y = "Percent change (%)", x = "Date", title = "Percent change^ in 7 day rolling average"), data = ott_covid_waste)
```

\* 7-day averages are midpoint rolling averages, calculated using the results from previous and later days. The frequency of testing changes over time. Currently, Ottawa wastewater testing is performed daily.

\* New cases (reported date) is the date reported by the laboratory. New cases (episode date) is the earliest onset, test or reported date.

\* Viral copies data is normalized using a seasonally stable fecal biomarker; pepper mild mottle virus (PMMoV). See [methods](/model/) for more details.

^ Percent change in 7-day average is calculated by comparing the 7-day average (previous day 1 to 7) with a lagged 7-day average (days 8 to 14).

Data to `r as.character(tail(ott_covid_waste$date, n=1))`